Progress Report for CSS 586 Course Project: \\ Modeling Latent Patterns to Generate Symbolic Music

### 2021-05-05 papers/progress/alex/progress-alex.tex
playing with an LSTM-AE model for viola track

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Progress Report for CSS 586 Course Project: \\
Modeling Latent Patterns to Generate Symbolic Music}

@@

\begin{abstract}
  This report explores recent research in symbolic music modeling with
  deep learning and provides a progress report on a course project to
  train and evaluate generative models of classical and popular music
  on the MusicNet and Lakh MIDI datasets.
\end{abstract}

@@

Depending on the task, machine learning models of music may be trained
on the audio signal itself, either in a time domain or a frequency
domain representation, or they may be trained on a digital symbolic
representation of music, the most common of which is MIDI (Musical
Instrument Digital Interface) notation. MIDI is an encoding of music
as streams of bytes in one or more tracks or channels, each
representing a sequence of 128 possible pitch values (where 0 is the
lowest and 127 is the highest), along with timing, pressure and
instrument identifier values. A music transcription model may
transcribe an audio signal as a MIDI score, which can easily be
converted into other symbolic representations such as sheet music for
human performers to read from, while a synthesizer model can convert
MIDI representations into audio signals. A generative music model can
be either trained to generate raw audio, or trained to produce a
symbolic score that must be played by either a synthesizer or by
humans to produce an audio music performance. This project focuses on
generative modeling of symbolic (MIDI) music to generate original
musical compositions.

@@

MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}. This model is capable of generating
new melodies that blend two given melodies via latent space
interpolation.

@@

A major advantage of working with the symbolic representation of music
is that it is of far lower dimensionality than the raw audio waveforms
of a recorded performance, which makes it less computationally
expensive. However, there are many aspects of musical performance
style that are not captured by a symbolic representation, so the
expressiveness of symbolic generative models is limited
\cite{manzelli_conditioning_2018}.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which I plan to use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}. Including this second dataset may
help us to generalize the modeling approach beyond European classical
music to include other popular genres and associated instruments.

@@

In order to accommodate polyphonic music, I will convert each MIDI
file into a pianoroll representation, as visualized in Figure
\ref{pianoroll}, wherein each instrument track is a sparse matrix that
multi-hot encodes the velocity values for each of 128 possible pitch
levels at each timestep. To reduce dimensionality, I will clip the
note pitch values to a narrower range, excluding the rarely used notes
in the extreme high and low octaves, and encode chords (combinations
of notes played simultaneously on the same instrument) as distinct
tokens, so that

@@

Evaluation of generative models is challenging because there is no
equivalent of an accuracy metric like what is used in supervised
learning. For autoencoder models we can measure how accurately the
model can reconstruct its own inputs, but this does not tell us the
quality of the interpolated examples. Generative models are typically
evaluated using a combination of qualitative metrics whereby human
judges rate the quality of the generated examples (essentially a
Turing test), and quantitative metrics that assess the differences in
the parametric distributions of generated and real examples. Yang and
Lerch (2020) proposes a set of metrics informed by music theory, for
probabilistically evaluating how similar the generations are to known
sample distributions of real music \cite{yang_evaluation_2020}. These
metrics include counts, ranges, histograms and transition matrices of
pitches and note lengths, then the Kullback-Leibler divergence and
overlapping area of the probability density functions are used to
compare against known reference distributions per musical genre
\cite{yang_evaluation_2020}. Due to the cost and time requirements
associated with designing a human subjects experiment, we plan to
utilize this quantitative approach to the quality assessment of
generated samples.

@@

At this stage, I have decided on MIDI generation as the modeling
problem, downloaded several MIDI datasets, and begun exploring them to
learn how to interpret, process, and visualize them.

I have implemented several pianoroll data processing functions that
crop sections of MIDI files into fixed-length training examples
consisting of multi-track note sequences in symbolic time (beats) at a
given time resolution. I have started implementing a training data
processing pipeline that will read MIDI files from disk and transform
them into the pianoroll-style tensor representation, providing batches
of training examples to Keras model layers for fitting.

I have started designing a simple LSTM autoencoder to model short
(e.g. 16 beats or 4 bars) clips of monophonic (single instrument
track) music, in order to learn the basics of sequence model
architecture. I will create a sequence predictor model with a similar
architecture and compare it to the autoencoder. Next I will add
support for polyphonic music with 3-5 instrument tracks. Once
polyphonic music generation is working well, I plan to increase the
length of the sequences and make adjustments to the model architecture
to permit it to model longer phrases and paragraphs of music.

I aim to build upon the hierarchical decoding approach taken in the
MusicVAE \cite{roberts_hierarchical_2018} and MuseGAN
\cite{dong2017musegan} papers, possibly adding more than two recurrent
layers in the decoder, to address a key challenge in generative music
modeling: the ability to learn the layered patterns in music to
produce compositions with convincing long-term structure.

>>>>

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}
\copyrightyear{2021}
\acmYear{2021}
\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Progress Report for CSS 586 Course Project: \\
Modeling Latent Patterns to Generate Symbolic Music}

@@

\begin{abstract}
  This report explores recent research in symbolic music modeling with
  deep learning and provides a progress report on a course project to
  train and evaluate generative models of classical and popular music.
\end{abstract}

@@

Depending on the task, machine learning models of music may be trained
on the audio signal itself, either in a time domain or a frequency
domain representation, or they may be trained on a digital symbolic
representation of music, the most common of which is MIDI (Musical
Instrument Digital Interface) notation. MIDI is an encoding of music
as streams of bytes in one or more tracks or channels, each
representing a sequence of 128 possible pitch values (where 0 is the
lowest and 127 is the highest), along with timing, pressure and
instrument identifier values. A music transcription model may
transcribe an audio signal as a MIDI score, which can easily be
converted into other symbolic representations such as sheet music for
human performers to read from, while a synthesizer model can convert
MIDI representations into audio signals. A generative music model can
be trained either to generate raw audio, or to produce a symbolic
score that must be played by a synthesizer or by humans to produce an
audio music performance. This project focuses on the latter type of
modeling: generative modeling of symbolic (MIDI) music to compose
original musical scores.

@@

MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}. This model is trained on 16-bar
paragraphs of music and is capable of generating new melodies that
blend two given melodies via latent space interpolation.

@@

A major advantage of working with the symbolic representation of music
is that it is of far lower dimensionality than the raw audio waveforms
of a recorded performance, which makes it less computationally
expensive. However, there are many stylistic aspects of musical
performance that are not captured by a symbolic representation, and
may be specific to a particular performer, so the expressiveness of
symbolic generative models is limited in comparison
\cite{manzelli_conditioning_2018}.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which I plan to use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}. Including this second dataset may
help to generalize the modeling approach beyond European classical
music to include other popular genres and associated instruments.

@@

In order to accommodate polyphonic music, I will convert each MIDI
file into a pianoroll representation, as visualized in Figure
\ref{pianoroll}, wherein each instrument track is a sparse matrix that
multi-hot encodes the velocity values for each of 128 possible pitch
levels at each timestep. To reduce dimensionality, I will clip the
note pitch values to a narrower range, excluding the rarely used notes
in the extreme high and low octaves, and encode chords (combinations
of notes played simultaneously on the same instrument) as distinct
tokens, so that they can be one-hot encoded

@@

Evaluation of generative models is challenging because there is no
equivalent of an accuracy metric like what is used in supervised
learning. For autoencoder models we can measure how accurately the
model can reconstruct its own inputs, but this does not tell us the
quality of the interpolated examples. Generative models are typically
evaluated using a combination of qualitative metrics whereby human
judges rate the quality of the generated examples (essentially a
Turing test), and quantitative metrics that assess the differences in
the parametric distributions of generated and real examples. Yang and
Lerch (2020) proposes a set of metrics informed by music theory, for
probabilistically evaluating how similar the generations are to known
sample distributions of real music \cite{yang_evaluation_2020}. These
metrics include counts, ranges, histograms and transition matrices of
pitches and note lengths, then the Kullback-Leibler divergence and
overlapping area of the probability density functions are used to
compare against known reference distributions per musical genre
\cite{yang_evaluation_2020}. Due to the cost and time requirements
associated with designing a human subjects experiment, I plan to
utilize this quantitative approach to the quality assessment of
generated samples.

@@

At this stage, I have decided on MIDI generation as the modeling
problem, downloaded several MIDI datasets, and begun exploring them to
learn how to interpret, process, and visualize them. Due to the
complexity of the MIDI format and the multiple options for
representing it in memory, data representation has been a
time-consuming portion of the work in this project.

So far I have implemented in Python several pianoroll data processing
functions that crop sections of MIDI files into fixed-length training
examples consisting of multi-track note sequences in symbolic time
(beats) at a given time resolution. I have also started implementing a
training data processing pipeline that will read MIDI files from disk
and transform them into the pianoroll-style tensor representation,
providing batches of training examples to Keras model layers for
fitting.

I have begun designing a simple LSTM autoencoder to model short
(e.g. 16 beats or 4 bars) clips of monophonic (single instrument
track) music, in order to learn the basics of sequence model
architecture and determine how complex the model must be in order to
be capable of reconstructing music note sequences. I will create a
sequence predictor model with a similar structure and compare its
generative capability to that of the autoencoder. Next I will add
support for polyphonic music with 3-5 instrument tracks. Once
polyphonic music generation is working well, I plan to increase the
length of the sequences and make adjustments to the model architecture
to permit it to model longer phrases and paragraphs of music. I aim to
improve upon the hierarchical decoding approach taken in the MusicVAE
\cite{roberts_hierarchical_2018} and MuseGAN \cite{dong2017musegan}
papers, possibly adding more than two recurrent layers in the decoder
and providing even longer training examples, to address a key
challenge in generative music modeling: the ability to learn from the
layered patterns in music to produce compositions with convincing
long-term structure. A stretch goal that would significantly
differentiate my work is to produce the first model capable of
generating not just separate phrases of symbolic music, but a coherent
song from beginning to end.
